NumPy---Array Computation for Python

### 2019-12-20 body.tex
Minor edits
Updates to abstract and intro

NumPy is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Since its first stable release (1.0) in October 2006 it has unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
Not surprisingly, many important scientific research projects and
results rely on NumPy.
%[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
%  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
%  IDL is now below Python, Pangeo <-> xarray, Earthcube,
%  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
%  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
%  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]

NumPy consists of the \code{ndarray}-object along with utility functions to
operate on it.
Because of its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy array is
the {\it de facto} exchange format for array data in Python.
NumPy, while having been wildly successful as a library and array exchange
project, has now also become a standard language for expressing array and
tensor computations.  
It has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous, and is
implemented by, for example, popular tensor computation libraries.

@@

Before NumPy, two different Python array packages—Numeric and NumArray—existed.
The first began in the mid-90s and was developed by Jim Fulton, Jim Hugunin,
Paul Dubois, Konrad Hinsen, Travis Oliphant, and others, and was aimed at fast
manipulation of many small arrays.
By the beginning of the millennium, several libraries built on top of Numeric 
the provided fundamental scientific routines `were combined to form SciPy.
Due to the poor performance of Numeric on large data sets, NumArray was
developed at the Space Telescope Science Institute (STScI) to handle
large astronomical images such as those produced by the Hubble Space Telescope.
It was initially considered the successor to Numeric.
However, it performed poorly on small arrays.
This divide posed a challenge for SciPy as it wasn't clear how to support
both libraries.
Working with the NumArray and SciPy community, the maintainer of Numeric,
Travis Oliphant, started a major rewrite that aimed to be a ``best of both
worlds'' unification of the two projects.  This became known as NumPy.

@@

% BIDS -- UCB
% https://www.moore.org/grant-detail?grantId=GBMF5447
% $645,020 in 2016
% https://sloan.org/grant-detail/8222
% $659,359 in 2017
% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
In 2017, NumPy received its first large grants totaling 1.3M USD from the
Gordon \& Betty Moore and the Alfred P. Sloan foundations.
Stéfan van der Walt is the PI and manages four programmers working on the project.
These two grants focus on addressing the technical debt accrued over the years and
on setting in place standards and architecture to encourage more sustainable development.

>>>>

NumPy is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Since its first stable release (1.0) in October 2006 it has unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
Not surprisingly, many important scientific research projects and
results rely on NumPy.
It plays an essential role in many research analysis pipelines in
fields as diverse as physics, biology, astronomy, neuroscience,
material science, engineering, and chemistry.
% w/ citations
In particular, NumPy played a fundamental role in discovering
gravitational waves\cite{pycbc}, imaging black
holes\cite{eht-imaging}, and exploring the solar system (... Kepler? https://ui.adsabs.harvard.edu/abs/2012ascl.soft08004S/abstract).

...
%[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
%  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
%  IDL is now below Python, Pangeo <-> xarray, Earthcube,
%  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
%  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
%  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]

NumPy consists of the \code{ndarray}-object along with utility functions that
operate on it.
Because of its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy array is
the {\it de facto} exchange format for array data in Python.
NumPy, having been wildly successful as a library and array exchange
project, now also has become a standard language for expressing array and
tensor computations.
It has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous, and is
implemented by, among others, popular tensor computation libraries\cite{pytorch}.

@@

Before NumPy, two different Python array packages—Numeric and NumArray—existed.
The first began in the mid-90s and was developed by Jim Fulton, Jim Hugunin,
Paul Dubois, Konrad Hinsen, Travis Oliphant, and others, and was aimed at fast
manipulation of many small arrays.
By the beginning of the millennium, several libraries built on top of Numeric 
the provided fundamental scientific routines were combined to form SciPy.
Due to the poor performance of Numeric on large data sets, NumArray was
developed at the Space Telescope Science Institute (STScI) to handle
large astronomical images such as those produced by the Hubble Space Telescope.
It was initially considered the successor to Numeric.
However, it performed poorly on small arrays.
This divide posed a challenge for SciPy as it wasn't clear how to support
both libraries.
Working with the NumArray and SciPy community, the maintainer of Numeric,
Travis Oliphant, started a major rewrite that aimed to be a ``best of both
worlds'' unification of the two projects.  %This became known as NumPy.
This merger of Numeric and NumArray was called NumPy.

@@

% BIDS -- UCB
% https://www.moore.org/grant-detail?grantId=GBMF5447
% $645,020 in 2016
% https://sloan.org/grant-detail/8222
% $659,359 in 2017
% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
In 2017, NumPy received its first large grants totaling 1.3M USD from the
Gordon \& Betty Moore and the Alfred P. Sloan foundations.
Stéfan van der Walt is the PI and manages four programmers working on the project.
These two grants focus on addressing the technical debt accrued over the years and
on setting in place standards and architecture to encourage more sustainable development.
% http://doi.org/10.5281/zenodo.3585761
% http://doi.org/10.5281/zenodo.3585767

### 2019-12-20 paper.tex
Revise intro
Add SJvdW to NumPy grant funding
Updates to abstract and intro

\begin{abstract}
  NumPy is the fundamental array computing library for Python.
  It has been actively developed for over two decades and has
  more than 850 contributors, nearly 300,000 dependent repositories
  on GitHub, and millions of downloads per year.
  It is extensively used in research and teaching.
  According to a 2018 survey, NumPy is the second most commonly used machine
  learning tool among enterprises \cite{451report2018}.
  In this article, we provide an overview of the capabilities and development
  practices of the NumPy library and highlight some recent technical
  developments.
\end{abstract}

@@

S.J.v.d.W and K.J.M. were funded in part by the Gordon and Betty Moore
Foundation through Grant GBMF3834 and by the Alfred P. Sloan Foundation through
Grant 2013-10-27 to the University of California, Berkeley.
S.B., T.J.R., and W.W. were funded in part by the Gordon and Betty Moore 
Foundation through Grant GBMF5447 and by the Alfred P. Sloan Foundation through
Grant G-2017-9960 to the University of California, Berkeley.

@@

%C.R.H and T.E.O.
S.J.v.d.W and K.J.M. composed the manuscript with input from others.
All authors have contributed significant code, documentation, and/or expertise
to the NumPy project.
All authors reviewed the manuscript.

>>>>

\begin{abstract}
  NumPy is the fundamental array computing library for Python.
  It has been actively developed for over two decades and has
  more than 850 contributors, nearly 300,000 dependent repositories
  on GitHub, and millions of downloads per year.
  It is extensively used in research and teaching.
  According to a 2018 survey, NumPy is the second most commonly used machine
  learning tool among enterprises \cite{451report2018}.
  NumPy continues to evolve to meet the needs of its community, to
  remain relevant in an ever-changing computing landscape, and to be a
  fundamental pillar of support for the next decade of data science in
  Python.
  In this article, we provide an overview of the capabilities and development
  practices of the library and highlight some recent technical
  developments.
\end{abstract}

@@

S.J.v.d.W and K.J.M. were funded in part by the Gordon and Betty Moore
Foundation through Grant GBMF3834 and by the Alfred P. Sloan Foundation through
Grant 2013-10-27 to the University of California, Berkeley.
S.J.v.d.W, S.B., T.J.R., and W.W. were funded in part by the Gordon
and Betty Moore Foundation through Grant GBMF5447 and by the Alfred
P. Sloan Foundation through Grant G-2017-9960 to the University of
California, Berkeley.

@@

%C.R.H and T.E.O.
S.J.v.d.W and K.J.M. composed the manuscript with input from others.
S.B., K.S., T.J.R., and W.W. drafted sections.
All authors have contributed significant code, documentation, and/or expertise
to the NumPy project.
All authors reviewed the manuscript.
